---
title: 'Lab 1: Getting Started with RStudio and Descriptive Statistics'
output:
  pdf_document: default
  html_document:
    default: yes
    number_sections: yes
  word_document: default
---

R is an open source tool for statistical computing and graphics, supported by the R Foundation for Statistical Computing. This document has been adapted from:  https://acadgild.com/blog/introduction-to-r-a-beginners-guide/

# Start RStudio and Get Your Bearings

To start RStudio in the lab environment, follow your professor's instructions. You can also download R (cran.r-project.org) and RStudio (www.rstudio.com/products/rstudio/download/) onto your own computer so you will always have the power of R at your fingertips.

Notice the *Console* pane (left side of the application window).  This is where you type commands and see results.

In RStudio you enter commands in two ways:

- type the command directly in the Console pane and hit `<enter>`
- use the menus and buttons in the Files/Plots/etc. pane to generate the command automatically

Other components in the RStudio user interface:

- *Workspace* pane shows all the active objects in your current session.
- *Files* pane shows all the files and folders in your default workspace.
- *Plots* pane shows all your graphs.
- *Packages* pane lists packages or add-ons needed to run certain processes.
- *Help* pane gives you access to R documentation for additional info on a command.

    **Help tip**: In the Console, use a prepended `? to pull up the R documentation for any R command.

TRY THIS (enter in Console) to pull up R documentation on the `print` command:

```{r, echo=TRUE, results="hide"}
?print
```

# Scientific Notation

To specify a large number such as 3.2 million, we generally use scientific notation. The multiplier goes first, then an `e` to separate the multipler from the exponent, then the actual exponent (desired power of 10). For example, since 1 million is `10^6`, we need `e6` to express the idea of 1 million.

Scientific notation is also helpful for indicating small values, such as 0.000007 (7 millionths), using negative exponents.

```{r, echo=FALSE, results="hide", warning=FALSE, message=FALSE}
library(hanoverbase)
```

Examples (try these yourself):
```{r, echo=TRUE, results="hide"}
3.2e6   #    3.2 million
-8.7e9  #   -8.7 billion
1e-1    #    0.1
7e-6    #    0.000007
```

# Working with the Console

The RStudio Console includes a variety of features intended to make working with R more productive and straightforward. Here we describe two of the most popular features.

## Tab Completion

In the Console, you might begin typing a command and then be unsure if you're spelling it correctly. Type the initial part of the command and then hit `<tab>`, and RStudio will show you all commands which begin in that way. Sometimes the completion list appears before you hit `<tab>`.

For example, to see the commands which start with "set", type "set" in the Console and hit `<tab>`. In the resulting popup, scroll through the "set" commands.  When your cursor hovers over one of them, a helpful synopsis of the command appears.

## Command History

As you work with R, you'll often want to re-execute a command which you previously entered. The RStudio Console supports the ability to recall previous commands using the History pane and the arrow keys. The History pane is typically situated at the upper right of the window. Click the History pane to see all the commands you have entered in the current session. Double-clicking any of these will paste it into the Console.

Alternatively, you can use the up/down arrow keys while in the Console, to scroll through the history.

A common workflow is to incrementally build a task by starting with a simple command, then using the up arrow to recall it and extend it by adding more steps.

## Variable Assignments

Variable assignments allow us to store the results of complicated commands so they can easily be recalled in the future. To assign a value to a name (or "variable"), use `<-` or `=`. When you make an assignment, you do not see a result right away, but a value has been stored for future use.

Try the following and observe the results:

```{r, results="hide"}
x <- 5
y <- 3
x + y
x <- 10
z <- x + y
z
```

If you have written the above correctly, the *Environment* pane will will show the current values of `x`, `y` and `z`.

# Formulas

We will use "formulas" in our graphing commands, to specify the variable(s) to display and any conditioning variables. Here are some examples (these will make more sense in future sections):

- to display variable x alone we would use the formula:  `~x`
- to display the numeric variable x for each level of factor A: `~x | A`
- to display the relationship between numeric variables y and x (treating y as the response): `y~x`

# The Piping Command (`%>%`)

You will see the "pipe" operator, %>%, in some of the commands below. The pipe operator takes the left-hand side (LHS) of the pipe and uses it as the first argument of the function on the right-hand side (RHS) of the pipe. As a very basic example, `1:10` is the sequence of integers from 1 to 10 inclusive. What is the sum of this sequence? What is the median?

TRY THIS:
```{r, results="hide"}
1:10               # the sequence 1 to 10
1:10 %>% sum()     # sum of the sequence 1 to 10
sum(1:10)          # The same thing
1:10 %>% median()  # median of the sequence 1 to 10
```

# Load Data and Start your Investigation

The U.S. 2010 Census generated a wealth of data. We will see later how to download data from the web and other sources. For now, we will use a built-in dataset, containing information about the U.S. counties.

To load the data set, type:
```{r}
data(counties)
```

In the Environment pane (upper right), you should now see a `counties` entry under Data. Click on that entry to view the data in its own pane. This the same as using the `View` command:

```{r, echo=TRUE, eval=FALSE}
View(counties)
```
     
Notice that you can sort the data file according to any column by clicking the arrows which are just to the right of each column heading. **Warning**: The `NA` (not available) answers will always sort at the bottom, whether you sort ascending or descending. Use scrolling and sorting to answer the following questions:

\newpage

1. How many counties are there in the U.S.?

\vfill

2. Name the top 3 counties by population (2010); give the county name, the state name and the population. For each of these three counties, explain why the county has such a large population. (Use internet search as needed.)

\vfill
\vfill
\vfill
\vfill


3. List the 3 counties with the least population (2010), including their populations. Use the internet to find an interesting fact about each county.

\vfill
\vfill
\vfill
\vfill

\newpage

# Histogram and Summary Statistics, One Quantitative Variable

The most popular graph for showing the distribution of a single quantitative variable is the histogram. Read about the histogram command, which is part of the Lattice package:

TODO: Replace with something more straightforward
```{r}
?lattice     
```

In the documentation, follow the link for "histogram".  Scroll to the end to see examples.

The following will draw the default histogram for pop2010:

```{r, fig.show="hide"}
histogram(~pop2010, data = counties)
```

You can get a bigger version of the graph by clicking the Zoom button in the **Plots** pane.

4. Describe the histogram (shape, center, spread, peak(s), outliers).

\vfill

This view of the pop2010 variable is unsatisfying. Almost all of the counties are thrown into a single bin, resulting in very little detail being visible. To get a better look at the distribution, we note that there are some extremely high outliers. We can filter the data to exclude counties with population greater than 2 million (require pop2010 <= 2e6). To accomplish this we will use the "piping" command to pipe the counties through a `filter`. The command for this is:
```{r, fig.show="hide"}
histogram(~pop2010, data = counties %>% filter(pop2010 <= 2e6))
```

5. This histogram is still lacking in detail, and we should filter out even more values. Adjust the histogram command to use a lower cutoff point than 2 million. You will need to use your judgment to find a cutoff so that the main pattern of the distribution is clearly shown. What is the mode of the distribution (this represents a "typical" county population)? How does that value make sense given your understanding of U.S. counties?

\vfill



